---
title: "Sample points from multiple DEMs"
output:
  html_document: 
    toc: true
  html_notebook:
    df_print: paged
params:
  output_dir:
    label: Output Directory
    value: c:/tempDir/out/
    input: text
  data_dir:
    label: Data Directory
    value: c:/tempDir/in
    input: text
  length_scale:
    label: Length Scale (m)
    value: 15.0
    input: numeric
  input_list:
    label: Text file containing a list of DEM .flt rasters and initiation point shapefiles.
    value: c:/tempDir/in/DEMs/input_list.txt
    input: text
  inner_buffer:
    label: Inner Buffer for Landslide Points (m)
    value: 15.0
    input: numeric
  outer_buffer:
    label: Outer Buffer for Landslide Points (m)
    value: 250.0
    input: numeric
  expansion_factor:
    label: Expansion Factor
    value: 1.1
    input: numeric
  duration_list:
    label: List of storm durations
    value: duration_list.txt
    input: text
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE
)
```

```{r load, include = FALSE}
library(TerrainWorksUtils)
library(terra)
library(stringr)
library(shiny)
library(car)
```

Julia has written a function "dem_to_model" that works with multiple DEMs, but I'd like to separate the steps explicitly so we can see what's being done and look at intermediate results. I've used the Fortran "watershed" program to clip out small basins containing landslide initiation points and runout tracks from the DOGAMI Special Paper 53 landslide inventory. These are output as binary floating point files (.flt) along with point shapefiles of the initiation points within each small basin. 

The names for these raster and point shapefiles are listed in a text file specified in the parameter list for this notebook. The directory containing the raster and vector files must include projection (.prj) files. The input file has a header line "DEM, Initiation_points" followed by one row for each input DEM and its corresponding initiation-point shapefile, separated by a comma. Use full-path names; this allows files from different data directories to be used. The rasters created by the code chunks below are written to the output directory specified in the parameter list. They are not necessarily deleted after completing the code chunk that creates them because that same chunk first looks to see if they exist. When experimenting or debugging, it saves time to not build them each time. However, if you change the input DEMs, you need to delete any files in the output directory; otherwise, they'll get mistakenly read in.

For now, I'll stick with the same set of explanatory variables explored with PFA_landslide_initiation.Rmd: gradient, mean curvature, and partial contributing area for a range of storm durations. A list of storm durations can be specified using the "duration_list" input parameter; the defaults are 6, 12, 24, and 48 hours.

```{r set_up_covariates}
output_dir <- params$output_dir

if (!dir.exists(output_dir)) {
  dir.create(output_dir)
}

length_scale = params$length_scale # Default is a 15-meter length scale (7.5-meter radius) for calculating derivatives.

data_files <- read.table(params$input_list, 
                         header = TRUE,
                         sep = ",")

```

Fist iterate through the DEMs, building covariate rasters for each and overlaying the landslide initiation-point locations to get the covariate values associated with each landslide. I'll store the accumulated values in dataframe ls_pnts. Let's take a quick look at the size of the input data files:

```{r data_file_size}
  num_files <- nrow(data_files)
  num_points <- 0
  for (i in 1:num_files) {
    initiation_points <- terra::vect(paste0(str_squish(data_files$Initiation_points[[i]]), ".shp"))
    num_points <- num_points + nrow(initiation_points)
  }
```

This list contains `r print(num_files)` files with `r print(num_points)` landslide initiation points.

``` {r landslide_point_values, results = 'hide'}
ls_pnts <- data.frame()

for (i in 1:num_files) {
  DEM <- paste0(data_files[i,1], ".flt")
  initiation_points <- terra::vect(paste0(str_trim(data_files[i,2]), ".shp"))
  
# Binary floating point rasters are created in the output_dir by function TerrainWorksUtils::elev_deriv. 
# It can take a long time to create them, so if they're already there, read them in.
  grad_name = paste0("gradient_", as.character(i))
  mean_name = paste0("meancurv_", as.character(i))
  
  if (!(file.exists(paste0(output_dir, grad_name, ".flt")) & file.exists(paste0(output_dir, mean_name, ".flt")))) {
    
    rasters <- c(paste0("GRADIENT,", output_dir, grad_name),
                 paste0("MEAN CURVATURE,", output_dir, mean_name))

    TerrainWorksUtils::elev_deriv(rasters = rasters,
                                  length_scale = length_scale,
                                  dem = DEM,
                                  scratch_dir = output_dir)
    
  }

  derivs <- terra::rast(paste0(output_dir, grad_name, ".flt"))
  names(derivs) <- "gradient"
  derivs <- c(derivs, terra::rast(paste0(output_dir, mean_name, ".flt")))
  names(derivs[[2]]) <- "mean_curv"
  
  # Get the gradient and mean curvature values at each initiation site and append them to the ls_pnts dataframe
  vals <- terra::extract(derivs,
                         initiation_points,
                         method = "bilinear")
  
  # Add the x-y coordinates of each point
  coords <- terra::crds(initiation_points)
  vals$x <- coords[,1]
  vals$y <- coords[,2]
  
  ls_pnts <- rbind(ls_pnts,  vals)
  
}

```

The ls_pnts dataframe provides the upper and lower limits for gradient and mean curvature in defining the analysis region from which to sample nonlandslide points. 

```{r range}
grad_min <- min(ls_pnts$gradient)
grad_max <- max(ls_pnts$gradient)
mean_min <- min(ls_pnts$mean_curv)
mean_max <- max(ls_pnts$mean_curv)

if (!params$expansion_factor == 1) {
  range <- grad_max - grad_min
  dgrad <- (params$expansion_factor - 1.0) * (range / 2.)
  grad_min <- grad_min - dgrad
  grad_max <- grad_max + dgrad
  range <- mean_max - mean_min
  dmean <- (params$expansion_factor - 1.0) * (range / 2.)
  mean_min <- mean_min - dmean
  mean_max <- mean_max + dmean
  }

```

We could use the create_analysis_region_mask function here, but that works with a set of initiation_points associated with a single DEM. Rather, we need to set the analysis region for each DEM independently, but based on the range of gradient and mean curvature values over the entire set of DEMs. We got that range above; here we set the analysis region for each DEM. While iterating through the DEM files, we'll also build the partial-accumulation-area rasters and the the values for the landslide and nonlandslide points from these.

``` {r, results = 'hide'}
range_mx <- matrix(c(grad_min, mean_min, grad_max, mean_max), ncol = 2, nrow = 2)
colnames(range_mx) <- c("min", "max")
rownames(range_mx) <- c("gradient", "mean_curv")

nonls_pnts <- data.frame()
ls_pnts <- data.frame()

duration_list <- params$duration_list
if (file.exists(params$duration_list)) {
  durations <- read.csv(duration_list, header = FALSE)
} else {
  durations <- c(6, 12, 24, 48)
}

for (i in 1:num_files) {
  DEM <- paste0(data_files[i,1], ".flt")
  initiation_points <- terra::vect(paste0(str_trim(data_files[i,2]), ".shp"))
  
  grad_name = paste0("gradient_", as.character(i))
  mean_name = paste0("meancurv_", as.character(i))
  covariates <- terra::rast(paste0(output_dir, grad_name, ".flt"))
  names(covariates) <- "gradient"
  covariates <- c(covariates, terra::rast(paste0(output_dir, mean_name, ".flt")))
  names(covariates[[2]]) <- "mean_curv"
  
  # Get the partial contributing areas here; we don't need to save the rasters.
  # This does take awhile.
  
  k = 1.0 # saturated hydraulic conductivity in meters per hour

  for (j in 1:ncol(durations)) {
    d <- durations[[j]]
    raster_name <- paste0("pca_k1_", as.character(d))
    covariates <- c(covariates, 
                    TerrainWorksUtils::contributing_area(raster = paste0(output_dir, raster_name),
                                                         dem = DEM, 
                                                         length_scale = length_scale, 
                                                         k = k, 
                                                         d = d, 
                                                         scratch_dir = output_dir))
    names(covariates[[2+j]]) <- raster_name
    
  }
  
  vals <- terra::extract(covariates,
                         initiation_points,
                         method = "bilinear")
  
  coords <- terra::crds(initiation_points)
  vals$x <- coords[,1]
  vals$y <- coords[,2]
  
  ls_pnts <- rbind(ls_pnts, vals)
  
# Now sample the nonlandslide analysis zone
  neg_region <- TerrainWorksUtils::make_neg_region(positive_points = initiation_points, 
                                                   ref_raster = DEM, 
                                                   inner_buffer = params$inner_buffer, 
                                                   outer_buffer = params$outer_buffer,
                                                   return_raster = TRUE)

# Set values within the inner buffer and beyond the outer buffer to nodata
  analysis_mask <- TerrainWorksUtils::mask_by_range(covariates[[1:2]], range_mx)
  analysis_mask <- analysis_mask * neg_region
  
  covariates <- covariates * analysis_mask
  
  num_points <- ceiling(length(initiation_points))

  neg_points <- TerrainWorksUtils::sample_points(count = num_points,
                                                 region = analysis_mask,
                                                 buffer = FALSE,
                                                  rseed = 12)
  
  vals <- terra::extract(covariates,
                         neg_points,
                         method = "bilinear")
  
  coords <- crds(neg_points)
  vals$x <- coords[,1]
  vals$y <- coords[,2]
  
  nonls_pnts <- rbind(nonls_pnts, vals)
  
}

# save the sample points
ls_pnts$class <- "pos"
nonls_pnts$class <- "neg"
sample_pnts <- rbind(ls_pnts, nonls_pnts)
save(sample_pnts, file = paste0(output_dir, "sample.Rdata"))

```

Let's compare the landslide and nonlandslide points.
```{r compare_grad}

par(mfrow = c(1, 2))

ls_grad_den <- density(ls_pnts$gradient)

nonls_grad_den <- density(nonls_pnts$gradient)

ymax <- max(max(ls_grad_den$y), max(nonls_grad_den$y))

num_ls = length(ls_pnts)
num_nonls = length(nonls_pnts)

this_title = paste0("N nonlandslide = ", as.character(nrow(nonls_pnts)))

plot(nonls_grad_den,
     ylim = c(0., ymax),
     col = "black",
     lwd = 2,
     main = this_title,
     xlab = "Gradient")

lines(ls_grad_den,
      col = "red",
      lwd = 2)

grid(nx = NULL,
     ny = NULL,
     col = "gray",
     lty = "dotted",
     lwd = 1)

legend(x = "topleft",
       cex = 0.7,
       bty = "n",
       legend = c("Nonlandslide", "Landslide"),
       lty = c(1,1),
       col = c("black", "red"),
       lwd = 2)

ls_curv_den <- density(ls_pnts$mean_curv)

nonls_curv_den <- density(nonls_pnts$mean_curv)

ymax <- max(max(ls_curv_den$y), max(nonls_curv_den$y))

this_title = paste0("N landslide = ", as.character(nrow(ls_pnts)))

plot(nonls_curv_den,
     ylim = c(0., ymax),
     col = "black",
     lwd = 2,
     main = this_title,
     xlab = "Mean Curvature")

lines(ls_curv_den,
      col = "red",
      lwd = 2)

grid(nx = NULL,
     ny = NULL,
     col = "gray",
     lty = "dotted",
     lwd = 1)

legend(x = "topright",
       cex = 0.7,
       bty = "n",
       legend = c("Nonlandslide", "Landslide"),
       lty = c(1,1),
       col = c("black", "red"),
       lwd = 2)

```
```{r scatterplot}
scatterplot(mean_curv ~ gradient, ls_pnts)
```

```{r}
num_d = length(durations)
par(mfrow = c(ceiling(num_d / 2), 2))

for (i in 1:num_d) {

  d <- durations[[i]]
  this_col <- 3 + i
  vals <- data.frame(ls_pnts[this_col])
  ls_den <- density(vals[[1]])
  vals <- data.frame(nonls_pnts[this_col])
  nonls_den <- density(vals[[1]])

  ymax = max(max(ls_den$y), max(nonls_den$y))

  plot(ls_den,
       ylim = c(0., ymax),
       lwd = 2,
       col = "red",
       main = paste0("Duration ",d,"hrs"))
  
  lines(nonls_den,
        lwd = 2,
        col = "blue")

  legend(x = "topleft",
         cex = 0.7,
         legend = c("Sample", "Landslide"),
         lty = c(1,1),
         col = c("blue", "red"),
         lwd = 2,
         bty = "n")

}
```
```{r}
vals <- ls_pnts[,-1] # remove the ID column
vals <- vals[,-6:8] # remove class column
scatterplotMatrix(vals,
                  main = "Landslide Points")

vals <- nonls_pnts[,-1]
vals <- vals[,-6:8]
scatterplotMatrix(vals, 
                  main = "Nonlandslide Points")
```
